The X-linked intellectual disability gene KDM5C is a sex-biased brake against germline programs in somatic lineages

### 2024-10-29 submission/manuscript.Rmd
intro edits

A pivotal step in the evolution of multicellularity is the division labor among cellular lineages, including distinguishing the germline from the soma. In the early embryo, genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications. Failure to repress germline genes in somatic lineages is common signature of cancer and observed in select neurodevelopmental disorders, however it is currently unclear how factors like development and sex influence their repression and somatic misexpression. Here, we examine how cellular context influences the development of somatic tissue identity in mice with loss of lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, ovary, and testis genes within the brain. By developing a comprehensive list of mouse germline-enriched genes, we found late-stage spermatogenesis genes and not somatic testicular genes were highly enriched within the _Kdm5c_-KO brain. Germline genes are typically decommissioned in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Characterizing germline gene misexpression in males and female mutants revealed germline gene repression is sexually dimorphic, with female EpiLCs requiring a higher dose of KDM5C to maintain germline gene suppression. Although many germline mRNAs are expressed in the _Kdm5c_-KO brain and EpiLCs, KDM5C recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation during ESC to EpiLC differentiation. Late-stage spermatogenesis genes lacking KDM5C and devoid of promoter CGIs can become expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a brake against runaway activation of germline developmental programs in somatic lineages.

**notes**

@@

A single genome holds the instructions for generating the myriad of cell types found within an organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Human genetic studies revealed impaired chromatin regulation commonly occurs in both cancer[@zhaoLanguageChromatinModification2021; @chiCovalentHistoneModificationsmiswritten2010; @berdascoAberrantEpigeneticLandscape2010] and neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018]. While many studies have identified their importance for regulating tumor suppressor genes and brain-specific transcriptional programs, loss of chromatin regulators can also cause ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of testis genes in colon tumors[@feichtingerMetaanalysisClinicalData2012] or liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. However, the mechanisms driving ectopic gene expression and its impact upon cancer and neurodevelopment are still poorly understood.

Separation of germline and somatic cellular identity is a pivotal step in the evolution of multicellularity and sexual reproduction[XXX, @devlinPanmetazoanViewGermlinesoma2023]. In mammals, chromatin regulators decommission germline genes in somatic lineages when the early embryo transitions from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes, such as those involved in later stages of oogenesis and spermatogenesis, employ the same silencing mechanisms. Furthermore, because many studies have focused on the silencing of key marker genes during early male embryogenesis, much is unknown about how cellular context influences the manifestation of germline gene misexpression. Given that impaired soma-germline demarcation is a signature of aggressive cancers and observed in select neurodevelopmental disorders (NDDs)[XXX; @berdascoAberrantEpigeneticLandscape2010; @bonefasSomatogermlineTransformationChromatinlinked2021], elucidating the impact of factors like sex and development is crucial for determining the mechanisms governing long-term misexpression of germline genes.

Here, we employed a genome-wide analyses to explore the impact of cellular context upon loss of tissue identity in mice lacking the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C). KDM5C lies on the X chromosome and erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007]. Somatic loss of KDM5C promotes tumorigenicity in a variety of cancer types[@shenSuppressionEnhancerOveractivation2016; @chenE6ProteinExpressed2018; @zhengDeficiencyXinactivationEscaping2021], while germline mutations cause the NDD Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, social deficits, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017; @bonefasSexuallyDimorphicAlterations2023]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some testis germline genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, what mechanisms underlie their dysregulation, and how KDM5C interacts with other known germline silencing mechanisms.

@@

Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes unique to the testis[@scandagliaLossKdm5cCauses2017]. It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain. We thus systematically tested whether other tissue-specific genes are misexpressed in the male brain with constitutive knockout of _Kdm5c_ (_Kdm5c^-/y^_, 5CKO in figures)[@vallianatosMutuallySuppressiveRoles2020] by using a published list of mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017].

@@

Given that germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown. In _C. elegans_, somatic misexpression of germline genes via loss of _Retinoblastoma (Rb)_ homologs results in enhanced piRNA signaling and ectopic P granule formation in neurons[@wangSomaticMisexpressionGermline2005; @wuRepressionGermlineRNAi2012]. Ectopic testicular germline transcripts have also been observed in a variety of cancers, including brain turmors in _Drosophila_ and mammals[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and shown to promote cancer progression[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012; @ninBiologyCancerTestisAntigens2023]. Intriguingly, mouse models and human cells for other chromatin-linked NDDs also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], such as mutations in DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, and methyl-CpG -binding protein 2 (MECP2). Recently, the transcription factor ZMYM2 (ZNF198), whose mutation causes a NDD (OMIM #619522), was also shown to repress germline genes by promoting H3K4me removal and CpGme[@graham-paquinZMYM2EssentialMethylation2023]. Thus, KDM5C is among a growing cohort of neurodevelopmental disorders with erosion of the germline-soma boundary. Further research is required to determine the impact of these germline genes upon neuronal functions and the extent to which this phenomenon occurs in humans.

>>>>

The division labor among cellular lineages is a pivotal step in the evolution of multicellularity. In mammals, the soma-germline boundary is formed during early embryogenesis, when genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications at promoter CpG islands (CGI). Somatic misexpression of germline genes is a signature of cancer and observed in select neurodevelopmental disorders. However it is currently unclear if all germline genes use the same repressive mechanisms and if factors like development and sex influence their dysregulation. Here, we examine how cellular context influences the development of somatic tissue identity in mice lacking lysine demethylase 5c (KDM5C), an X chromosome eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, and germline genes within the brain. By developing a comprehensive list of mouse germline-enriched genes, we found the _Kdm5c_-KO brain aberrantly expresses late-stage spermatogenesis genes while _Kdm5c_-KO epiblast-like cells (EpiLCs) primarily expressed key regulators of germline identity and meiosis, such as _Dazl_ and _Stra8_. KDM5C-mediated germline gene repression is sexually dimorphic, as female EpiLCs require a higher dose of KDM5C to maintain germline silencing. KDM5C represses germline genes during ESC to EpiLC differentiation, during which it is recruited to a subset of CGI-containing germline identity gene promoters to facilitate CpG methylation. However the majority of germline gene promoters did not harbor CGIs and were unbound by KDM5C, including late-stage spermatogenesis genes that are expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Altogether, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a sex-biased brake against runaway activation of germline developmental programs in somatic lineages.

**notes**
- How does sex and development influence germline gene repression?
- How does KDM5C interact with known silencing machinery? 
- Do all germline genes use teh same repressive mechanisms

  aberrantly expressed , including _Dazl_ and _Stra8_.
  - Germline genes are typically decommissioned in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.

@@

A single genome holds the instructions to generate the myriad of cell types found within an organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Human genetic studies revealed impaired chromatin regulation commonly occurs in both cancer[@zhaoLanguageChromatinModification2021; @chiCovalentHistoneModificationsmiswritten2010; @berdascoAberrantEpigeneticLandscape2010] and neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018]. While many studies have identified their importance for regulating tumor suppressor genes and brain-specific transcriptional programs, loss of chromatin regulators can also cause ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of testis genes in colon tumors[@feichtingerMetaanalysisClinicalData2012] or liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. However, the mechanisms driving ectopic gene expression and its impact upon cancer and neurodevelopment are still poorly understood.

Separation between germline and somatic cellular identity is a pivotal step in the evolution of multicellularity and sexual reproduction[@hanschenMulticellularityDrivesEvolution2018; @michodEvolutionaryTransitionsIndividuality2011; @chenMulticellularitySexHelped2021; @devlinPanmetazoanViewGermlinesoma2023]. In mammals, chromatin regulators decommission germline genes in somatic lineages when the early embryo transitions from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) at their CpG islands (CGIs) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes, such as those involved in the later stages of oogenesis and spermatogenesis, employ the same silencing mechanisms. Furthermore, because many studies have focused on the silencing of key marker genes during early male embryonic development, much is unknown about how cellular context (i.e. sex and tissue environment) influences the manifestation of germline gene misexpression. Intriguingly, impaired soma-germline demarcation is a signature of aggressive cancers and observed in select neurodevelopmental disorders (NDDs)[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017; @berdascoAberrantEpigeneticLandscape2010; @bonefasSomatogermlineTransformationChromatinlinked2021]. Thus, elucidating how cell context contributes to germline gene dysregulation will reveal novel mechanisms governing these pathologies.

Here, we employed genome-wide analyses to explore loss of tissue identity in mice lacking the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C). KDM5C lies on the X chromosome and erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007]. Somatic loss of KDM5C promotes tumorigenicity in a variety of cancer types[@shenSuppressionEnhancerOveractivation2016; @chenE6ProteinExpressed2018; @zhengDeficiencyXinactivationEscaping2021], while germline mutations cause the NDD Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, social deficits, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017; @bonefasSexuallyDimorphicAlterations2023]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some testis germline genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, what mechanisms underlie their dysregulation, and how KDM5C interacts with other known germline silencing mechanisms.

@@

Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes unique to the testis[@scandagliaLossKdm5cCauses2017]. It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain. We first systematically tested whether other tissue-specific genes are misexpressed in the male brain with constitutive knockout of _Kdm5c_ (_Kdm5c^-/y^_, 5CKO in figures)[@vallianatosMutuallySuppressiveRoles2020] by using a published list of mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017].

@@

Given that germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown. In _C. elegans_, somatic misexpression of germline genes via loss of _Retinoblastoma (Rb)_ homologs results in enhanced piRNA signaling and ectopic P granule formation in neurons[@wangSomaticMisexpressionGermline2005; @wuRepressionGermlineRNAi2012]. Ectopic testicular germline transcripts have also been observed in a variety of cancers, including brain turmors in _Drosophila_ and mammals and shown to promote cancer progression[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017; @janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012; @ninBiologyCancerTestisAntigens2023]. Intriguingly, mouse models and human cells for other chromatin-linked NDDs also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], such as mutations in DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, and methyl-CpG -binding protein 2 (MECP2). Recently, the transcription factor ZMYM2 (ZNF198), whose mutation causes a NDD (OMIM #619522), was also shown to repress germline genes by promoting H3K4me removal and CpGme[@graham-paquinZMYM2EssentialMethylation2023]. Thus, KDM5C is among a growing cohort of neurodevelopmental disorders with erosion of the germline-soma boundary. Further research is required to determine the impact of these germline genes upon neuronal functions and the extent to which this phenomenon occurs in humans.
